Cache Oblivious B-trees

### 2011-12-09 paper.tex
Adding updates to the background section.

Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
\cite{cobtree}. The paper presents three different data structures that make
different trade offs in the number of memory transfer they perform. However,
all of them achieve their performance by structuring their elements as a
strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
PMA play crucial roles in the other cache-oblivious data structures that were
developed after the CO B-tree and thus are described at length in the next two
sections.

@@

A nice consequence of using a vEB layout is that we know that it will use
$O(log_B N)$ memory transfers for searches [???, ???]. Thus, off the bat we
have achieve the desired number of memory transfers for search queries. It
still remains to be shown that the CO B-tree achieves the desired number of
memory transfers for dynamic inserts and deletes.

We know that the CO B-tree is going to use a balance search tree, but what
kind?. Strongly weight-balanced search trees play a key role in CO B-tree. The
weight of a node is $u$ is equal to the total number of $u$'s descendants plus
1. It can also be thought of as the sum of the weights of $u$'s children plus
1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.

Most weight-balanced search trees which requires that the heights of a node's
left and right subtrees are simply off by a constant factor. SWBST enforce the
stronger property that a node at height $h$ in the tree (height is the node's
distance from the leaves; leaves have a height of 1) has $\Theta c^h$
descendants where $c$ is some constant. This property becomes important later
on when laying the tree out in memory (see Section~\ref{}).

@@

\begin{equation}
\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}}
\end{equation}

@@

When we split a node, we naturally need to determine which children belong to
which of the two nodes. We cannot just divide the children evenly because it
may cause another violation. Thus, instead, we find the longest sequence of
elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
By doing this, we maintain the bound that $\frac{c^{h-1}{2} \leq w(u) \leq
2c^{h-1}}$. We cascade these changes up the tree until all of the ancestors
are balanced.

Deletions are supported in the exact same way, except, we check is a node has
a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
together and proceed up the tree. There is one edge case that needs to be
handled. After merging, you may produce a new node that has a weight larger
than the bounds. If this happens we immediately split the node again. Thus,
this can be thought of as stealing nodes from a neighbor.

WE have now shown that a SWBST achieves the desired property that the number
of descendants of a node is $c^h$. We can now show how to lay this tree out in
memory.

@@

Fist, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
transfers. Doing so may seem like attempting to achieve contradictory goals:
maintaining the locality of elements while keeping space available for new
inserts. The PMA uses a technique described in [???]. At a high level, the PMA
maintains a windows into the array. When a window becomes too full or too
empty, the elements are spread out within the context of a larger window. Two
parameters are required, the window size and the threshold for trigger the
spreading out. However, these parameters are not related to the block size of
our memory level and thus do not violate the CO model of computation. Overall,
the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
transfers.

Second, the PMA needs to handle the splitting and merging that will be caused
by the SWBST discussed in the previous section. THe PMA is able to accomplish
this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
fact that each node has $c^h$ descendants (recall, this is the key property
that SWBST provided). Thus, merging or splitting a node will cause $c^h$ nodes
to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
defitinot of SWBST in [???] that when a node is rebalanced, it will require
$c^h$ insertions or deletions until it is rebalanced again. Thus, the
amortized cost of each insertion and/or deletion is $O(1/B)$. When you split a
node, it may go all the way up the tree and thus overall it will incur $O((log
N) / B)$ memory transfers.

@@

Of these three terms, the dominating factor is maintain room for newly
inserted elements - $O((log^2N)/B)$. We know that SWBST achieve insertions and
deletitions in $O(log_B N)$ if they are cache-aware. Thus, adding PMA makes
the cache-oblivious and thereby incurring an additional $O((log^2N)/B)$ memory
transfers.

Therefore, overall, CO B-Tree achieve insertions and deletions in $O(log_B N +
(log^2N)/B)$ memory transfers. However, they still achieve the optimal
$O(log_B N)$ memory transfers for searches, matching the performance of
ordinary, cache-aware, B-trees.

>>>>

Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
\cite{cobtree}. The CO B-tree achieves its number of memory transfers by structuring its elements as a
strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
PMA play crucial roles in the other cache-oblivious data structures that were
developed after the CO B-tree and thus are described at length in the next two
sections.

@@

A nice consequence of using a vEB layout is that it is known that it will use
$O(log_B N)$ memory transfers for searches [???, ???]. Thus, the CO B-tree has already
achieved the desired number of memory transfers for search queries. It
still remains to be shown that the CO B-tree achieves the desired number of
memory transfers for dynamic inserts and deletes.

It is known that the CO B-tree is going to use a balance search tree, but what
kind? This is where the strongly weight-balanced search tree becomes useful. The
weight of a node is $u$ is equal to the total number of $u$'s descendants plus
1. It can also be thought of as the sum of the weights of $u$'s children plus
1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.

Most weight-balanced search trees which requires that the heights of a node's
left and right subtrees are simply off by a constant factor. The SWBST enforces the
stronger property that a node at height $h$ in the tree (height is the node's
distance from the leaves; leaves have a height of 1) has $\Theta c^h$
descendants where $c$ is some constant. This property becomes important later
on when laying the tree out in memory (see Section~\ref{sec:layout}).

@@

\begin{equation}
\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}
\end{equation}

@@

When we split a node, we naturally need to determine which children belong to
which of the two nodes. We cannot just divide the children evenly because it
may cause another violation. Thus, instead, we find the longest sequence of
elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
By doing this, we maintain the bound that $\frac{c^{h-1}}{2} \leq w(u) \leq
2c^{h-1}$. We cascade these changes up the tree until all of the ancestors
are balanced.

Deletions are supported in the exact same way, except, we check if a node has
a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
together and proceed up the tree. There is one edge case that needs to be
handled. After merging, you may produce a new node that has a weight larger
than the bounds. If this happens we immediately split the node again. Thus,
this can be thought of as stealing nodes from a neighbor.

We have now shown that the SWBST achieves the desired property that the number
of descendants of a node is $c^h$. We can now show how to lay this tree out in
memory.

@@

Fist, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
transfers. Doing so may seem like attempting to achieve contradictory goals:
maintaining the locality of elements while keeping space available for new
inserts. The PMA uses a technique described in [???]. At a high level, the PMA
maintains a windows into the array. When a window becomes too full or too
empty, the elements are spread out within the context of a larger window. Two
parameters are required, the window size and the threshold for trigger the
spreading out. However, these parameters are not related to the block size of
the memory level and thus do not violate the CO model of computation. Overall,
the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
transfers.

Second, the PMA needs to handle the splitting and merging that will be caused
by the SWBST discussed in the previous section. The PMA is able to accomplish
this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
fact that each node has $c^h$ descendants (recall, this is the key property
that SWBST provided). Thus, merging or splitting a node will cause $c^h$ nodes
to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
definition of SWBST in [???] that when a node is rebalanced, it will require
$c^h$ insertions or deletions until it is rebalanced again. Thus, the
amortized cost of each insertion and/or deletion is $O(1/B)$. When you split a
node, it may go all the way up the tree and thus overall it will incur $O((log
N) / B)$ memory transfers.

@@

Of these three terms, the dominating factor is maintaining room for newly
inserted elements - $O((log^2N)/B)$. We know that the SWBST achieves insertions and
deletions in $O(log_B N)$ if it is cache-aware. Thus, adding PMA makes
the SWBST cache-oblivious at the cost of incurring an additional $O((log^2N)/B)$ memory
transfers.

Therefore, overall, the CO B-Tree achieve insertions and deletions in $O(log_B N +
(log^2N)/B)$ memory transfers. However, it still achieves the optimal
$O(log_B N)$ memory transfers for searches, matching the performance of
ordinary, cache-aware, B-trees.
